---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# PHEstatmethods

This is an R package to support analysts in PHE to perform standard analysis using approved methodology.

This package is meant to remain up to date with approved methodology. Any feedback would be appreciated and can be provided using the Issues section of the GitLab repository, or by emailing PHDS@phe.gov.uk

## Installation

### From zip

Download this repository from GitLab and either build from source or do:

```{r source_pkg-installation, eval = FALSE}
source <- devtools:::source_pkg("C:/path/to/fingertips-master")
install(source)
```

### With devtools

You can install the latest version of PHEstatmethods from GitLab with:

```{r gl-installation, eval = FALSE}
if (!require(devtools)) install.packages("devtools")
devtools::install_git('https://gitlab.phe.gov.uk/Georgina.Anderson/PHEstatmethods', build_vignettes=TRUE)
```
