---
title: "Vignette for calculating DSRs using the PHEstatmethods package"
author: "Georgina Anderson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette for calculating DSRs for multiple geographies using data from the Data Lake}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# DRAFT - work-in-progress

## What does the sample code do?
This vignette documents the requirements and method for calculating DSRs using the phe_dsr function

The function can calculate DSRs for multiple geographic areas and/or multiple genders.  The function takes the following inputs:
* x - a numeric vector containing the numerators
* n 
* stdpop
* groupref


## Caution
* ensure that deaths data is padded so a record exists for every geography and age band combination - 
  even when there were no deaths in that group
* ensure the vectors x and n are ordered by the same area and ageband values
* ensure the standard population reference is ordered by the same age band values





# The following needs deleting (useful syntax from template):


## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
